---
title: "Functional ANOVA Example"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    keep_md: true
  pdf_document: default
---





```{r, include=FALSE}

###########################
## Load/Install Packages ##
###########################

pkgs = c("MASS","knitr","ggplot2","RColorBrewer","nlme","lme4","rjags")
for(p in pkgs) if(p %in% rownames(installed.packages()) == FALSE) {install.packages(p, dependencies = TRUE)}
for(p in pkgs) suppressPackageStartupMessages(library(p, quietly=TRUE, character.only=TRUE))
rm(p, pkgs)





#########################
## Set Overall Options ##
#########################

knitr::opts_chunk$set(fig.align="center", fig.width=8, fig.height=6)

```





<br>
<br>

## Simulate Three-Group One-Way Functional ANOVA Data

Three-group one-way functional ANOVA data were simulated using methods described in the paper "Fast Function-on-Scalar Regression with Penalized Basis Expansions" from Reiss et al:

$$
\begin{aligned}
y_{i}(t) &= \mu(t) + \beta_{gp(i)}(t) + \epsilon_{i}(t) \\
\mu(t) &= 0.4\arctan(10x - 5) + 0.6 \\
\beta_{1}(t) &= -0.5 e^{-10t} -0.04\sin(8t) - 0.3t + 0.5 \\
\beta_{2}(t) &= -(t - 0.5)^{2} - 0.15 \sin(13t) \\
\beta_{3}(t) &= -\beta_{1}(t) - \beta_{2}(t)
\end{aligned}
$$
where $t = m/200$ for $m=0,\dots,200$ and the error functions, $\epsilon_{i}(t)$, were simulated from a mean-zero Gaussian process with coveriance $V(s,t) = \sigma_{1} 0.15^{|s-t|} + \sigma_{2} \delta_{st}$ where $\delta_{st}=1$ if $s=t$ and $0$ otherwise. For the simulations, N=10 curves were simulated for each of the 3 groups, $\sigma_{1} = 0.15$, and $\sigma_{2} = 0.05$. So, we have `r 3*10*201` observations from `r 10*3` curves with 10 curves per group.





<br>
<br>

```{r}

###############################
## Create the "True" Dataset ##
###############################

## Model
## y_i(t) = mu(t) + b_gp(i)(t) + e_i(t) for t in [0,1]

## Values
t = 0:200/200     ## Evaluation time points
g = 3             ## Number of groups
N = 10            ## Number of curves per group, N = 10 or 60

## Coefficients from the paper
mu.t = 0.4*atan(10*t - 5) + 0.6
b1.t = -0.5*exp(-10*t) - 0.04*sin(8*t) - 0.3*t + 0.5
b2.t = -(t - 0.5)^2 - 0.15*sin(13*t)
b3.t = -b1.t - b2.t

## Mean functions by group
mean_1 = mu.t + b1.t
mean_2 = mu.t + b2.t
mean_3 = mu.t + b3.t

## Create the "true" dataset
true.data = data.frame(Y_True = c(mean_1, mean_2, mean_3),
                       Time = rep(t, 3),
                       Group = as.factor(c(rep(1, length(t)),
                                           rep(2, length(t)),
                                           rep(3, length(t)))) )

```





<br>
<br>

Note that the X variable, Time, was centered and scaled.
```{r}

#############################
## Generate Simulated Data ##
#############################

## Create a stacked dataframe for the simulated data
dat = data.frame(Y_True = c(rep(mean_1, N),
                            rep(mean_2, N),
                            rep(mean_3, N)),
                 Time = rep(t, N*g),
                 Group = as.factor( c(rep(1, N*length(t)),
                                      rep(2, N*length(t)),
                                      rep(3, N*length(t))) ),
                 ID = factor( rep(1:(N*g),
                                  length(t))[ order(rep(1:(N*g), length(t))) ] ) )
rm(mean_1, mean_2, mean_3)

## Define values
sigma_1 = 0.15    ## Simulations used 0.05 and 0.15
sigma_2 = 0.05    ## 0.05 for all simulations

## Calculate the covariance matrix
grid = expand.grid(t, t)
cov.matrix = apply(X = grid, MARGIN = 1,
                   FUN = function(x){ ((sigma_1^2)*0.15^(abs(x[1] - x[2]) )) +
                       (sigma_2^2)*as.numeric(x[1] == x[2]) } )
cov.matrix = matrix(data = cov.matrix, ncol = length(t), nrow = length(t))
rm(grid)

## Generate errors
seed = 87654321
set.seed(seed)
errors = mvrnorm(n = N*g, mu = rep(0, length(t)), Sigma = cov.matrix)

## Add to the dataframe
dat$Errors = c( errors[1:(N*g),] )

## Calculate simulated Y's
dat$Y_Sim = (dat$Y_True + dat$Errors)
rm(cov.matrix, errors, sigma_1, sigma_2)

## Pre-smooth data
dat$Y_Smooth = rep(NA, N*g*length(t))
for(i in 1:(N*g)){
  start = (i-1)*length(t) + 1
  end = i*length(t)
  dat$Y_Smooth[start:end] = with(dat,
                                 smooth.spline(Time[start:end], Y_Sim[start:end],
                                               nknots = 20))$y
}
rm(i, start, end, t, mu.t)

## Center and scale the time variable
dat$Time_Standardized = scale(dat$Time)

## Calculate higher-order time terms
dat$Time_Stan_2 = dat$Time_Standardized^2
dat$Time_Stan_3 = dat$Time_Standardized^3
dat$Time_Stan_4 = dat$Time_Standardized^4
dat$Time_Stan_5 = dat$Time_Standardized^5

```





```{r, include=FALSE}

## Save the data objects
save(list = ls(), file = "data/simulateddata.rda")

```




<br>
<br>
\pagebreak

#### Figure 1: The true effect functions along with the simulated data.
```{r, echo=FALSE}

## Create the plot
figure = ggplot() +
  geom_line(data = true.data, aes(x = Time, y = Y_True, color = Group), size = 1.25) +
  geom_point(data = dat, aes(x = Time, y = Y_Sim, color = Group), alpha = 0.3, size = 0.9) +
  theme_bw() +
  scale_color_brewer(palette = "Set1") +
  xlab("Time") +
  ylab("Y")

## Save the plot
ggsave(plot = figure,
       filename = "figure1.pdf",
       device = "pdf",
       path = "figs/",
       width = 8,
       height = 6)

## Return the plot
figure
rm(figure)

```





<br>
<br>
\pagebreak

#### Figure 2: The smoothed subject-specific curves from the simulated data.
```{r, echo=FALSE}

## Create the plot
figure = ggplot(data = dat, aes(x = Time, y = Y_Smooth, group = ID, color = Group)) +
  geom_line() +
  theme_bw() +
  scale_color_brewer(palette = "Set1") +
  xlab("Time") +
  ylab("Y")

## Save the plot
ggsave(plot = figure,
       filename = "figure2.pdf",
       device = "pdf",
       path = "figs/",
       width = 8,
       height = 6)

## Return the plot
figure
rm(figure)

```





<br>
<br>
\pagebreak

#### Figure 3: The group-level mean functions for the smoothed simulated data $\pm$ 2 point-wise standard deviations.
```{r, echo=FALSE}

## Find the mean heights at each time for each gender
group.means = aggregate(dat$Y_Smooth ~ dat$Time_Standardized + dat$Group, FUN = mean)
names(group.means) = c("Time","Group","Y")

## Find the gender-specific SD
group.sds = aggregate(dat$Y_Smooth ~ dat$Time_Standardized + dat$Group, FUN = sd)
names(group.sds) = c("Time","Group","SD")

## Combine means and sds
group.summ = merge(x = group.means, y = group.sds, by = c("Time","Group"), sort = FALSE)
rm(group.means, group.sds)

## Add 1 SD band
group.summ$Lower = group.summ$Y - 2*group.summ$SD
group.summ$Upper = group.summ$Y + 2*group.summ$SD

## Plot the gender-specific curves
figure = ggplot(data = group.summ, aes(x = Time, y = Y, fill = Group)) +
  geom_line() +
  theme_bw() +
  scale_color_brewer(palette = "Set1") +
  xlab("Time") +
  ylab("Y") +
  geom_ribbon(aes(ymin = Lower, ymax = Upper), alpha = 0.4) +
  scale_fill_brewer(palette = "Set1")

## Save the plot
ggsave(plot = figure,
       filename = "figure3.pdf",
       device = "pdf",
       path = "figs/",
       width = 8,
       height = 6)

## Return the plot
figure
rm(figure, group.summ)

```





<br>
<br>
\pagebreak

## Analyze Using lm()

Fit a 5th-order linear model with time-by-group and $\text{Time}^2$-by-group interactions using lm():
```{r}

## Fit the model
mod = lm(Y_Smooth ~ Group + Time_Standardized + Time_Stan_2 + Time_Stan_3 +
           Time_Stan_4 + Time_Stan_5 +
           Group*Time_Standardized + Group*Time_Stan_2, data = dat)

## Calculate the fitted values
dat$Y_Smooth_Fitted_lm = fitted(mod)

```





```{r, include=FALSE}

## Save the model object
save(mod, file = "output/model_lm.rda")
rm(mod)

```






<br>
<br>

#### Figure 4: The group-level mean functions from lm() using a 5th-order linear model with time-by-group and $\text{Time}^2$-by-group interactions.
```{r, echo=FALSE}

## Create the plot
figure = ggplot(data = dat, aes(x = Time, y = Y_Smooth_Fitted_lm, group = ID, color = Group)) +
  geom_line() +
  theme_bw() +
  scale_color_brewer(palette = "Set1") +
  xlab("Time") +
  ylab("Y")

## Save the plot
ggsave(plot = figure,
       filename = "figure4.pdf",
       device = "pdf",
       path = "figs/",
       width = 8,
       height = 6)

## Return the plot
figure
rm(figure)

```





<br>
<br>
\pagebreak

## Analyze Using Penalized Splines and lme()

Fit a penalized spline using lme():
```{r}

## Penalized basis functions
K = 24
qtiles = seq(0, 1, length = K + 2)[-c(1, K + 2)] 
knots = quantile(dat$Time_Standardized, qtiles)
random.basis = cbind(sapply(knots,
                            function(k){
                              I((dat$Time_Standardized - k)^2)*(dat$Time_Standardized - k > 0)
                              }
                            )
                     )
rm(qtiles)

## Define the fixed basis
fixed.basis = cbind(1,
                    as.numeric(dat$Group == "2"),
                    as.numeric(dat$Group == "3"),
                    dat$Time_Standardized,
                    dat$Time_Stan_2)

## Partition the random.basis matrix
random.basis.1 = as.numeric(dat$Group == "1")*random.basis
random.basis.2 = as.numeric(dat$Group == "2")*random.basis
random.basis.3 = as.numeric(dat$Group == "3")*random.basis

## Define Y
Y = dat$Y_Smooth

## Fit the model
group.1 = group.2 = group.3 = rep(1, nrow(dat))
mod = lme(Y ~ fixed.basis - 1, random = list(group.1 = pdIdent(~ random.basis.1 - 1),
                                             group.2 = pdIdent(~ random.basis.2 - 1),
                                             group.3 = pdIdent(~ random.basis.3 - 1)) )
rm(group.1, group.2, group.3)

## Extract the model coefficients
coefs = c( unlist(mod$coefficients$fixed), unlist(mod$coefficients$random) )

## Extract the lambda estimates
#var.e2 = mod$sigma^2
#var.u2 = getVarCov(mod)[1,1]

## Calculate the fitted values
dat$Y_Smooth_Fitted_lme = as.vector( cbind(fixed.basis,
                                           random.basis.1,
                                           random.basis.2,
                                           random.basis.3) %*% coefs )

## Remove R objects
rm(fixed.basis, random.basis, coefs, knots, Y)

```





```{r, include=FALSE}

## Save the model object
save(mod, file = "output/model_lme.rda")
rm(mod)

```





<br>
<br>
\pagebreak

#### Figure 5: The group-level mean functions using penalized splines and lme().
```{r, echo=FALSE}

## Create the plot
figure = ggplot(data = dat, aes(x = Time, y = Y_Smooth_Fitted_lme, group = ID, color = Group)) +
  geom_line() +
  theme_bw() +
  scale_color_brewer(palette = "Set1") +
  xlab("Time") +
  ylab("Y")

## Save the plot
ggsave(plot = figure,
       filename = "figure5.pdf",
       device = "pdf",
       path = "figs/",
       width = 8,
       height = 6)

## Return the plot
figure
rm(figure)

```





<br>
<br>
\pagebreak

## Analyze Using JAGS

Fit a 5th-order linear model with time-by-group and $\text{Time}^2$-by-group interactions using JAGS:
```{r}

## Define values
n = nrow(dat)
Y = dat$Y_Smooth
X = c(dat$Time_Standardized)
X_2 = c(dat$Time_Stan_2)
X_3 = c(dat$Time_Stan_3)
X_4 = c(dat$Time_Stan_4)
X_5 = c(dat$Time_Stan_5)
G.2 = as.numeric(dat$Group == "2")
G.3 = as.numeric(dat$Group == "3")

## Create the model
model_string <- "model{

# Likelihood
for(i in 1:n){
Y[i]   ~ dnorm(mu[i], inv.var)
mu[i] <- beta[1] + beta[2]*X[i] + beta[3]*X_2[i] + beta[4]*X_3[i] + beta[5]*X_4[i] + beta[6]*X_5[i] + beta[7]*G.2[i] + beta[8]*G.3[i] + beta[9]*X[i]*G.2[i] + beta[10]*X[i]*G.3[i] + beta[11]*X_2[i]*G.2[i] + beta[12]*X_2[i]*G.3[i]
}

# Prior for beta
for(j in 1:12){
beta[j] ~ dnorm(0,0.0001)
}

# Prior for the inverse variance
inv.var   ~ dgamma(0.01, 0.01)
sigma     <- 1/sqrt(inv.var)

}"

## Compile the model
model = jags.model(textConnection(model_string),
                   data = list(Y = Y,
                               n = n,
                               X = X,
                               X_2 = X_2,
                               X_3 = X_3,
                               X_4 = X_4,
                               X_5 = X_5,
                               G.2 = G.2,
                               G.3 = G.3),
                   n.chains = 3,
                   n.adapt  = 1000)

## Burn-in samples
update(model,
       1000,
       progress.bar = "none")

## Draw additional samples
samp = coda.samples(model,
                    variable.names = c("beta","sigma"),
                    thin = 3,
                    n.iter = 5000,
                    progress.bar = "none")

```





```{r, include=FALSE}

## Save the model objects
save(model,
     samp,
     file = "output/model_JAGS.rda")

```





<br>
<br>

Investigate the fit of the model and extract the fitted values:
```{r}

## Investigate the model
summary(samp)
plot(samp)

## Save the coefficients from the model
coefs = data.frame( summary(samp)$statistics )

## Extract the beta estimates
betas = coefs$Mean[ row.names(coefs) %in% paste0( paste0("beta[",1:12), "]") ]

## Calculate the fitted values
dat$Y_Smooth_Fitted_JAGS_5 = c(t(betas) %*% t(cbind(1, X, X_2, X_3, X_4, X_5,
                                                    G.2, G.3, X*G.2, X*G.3,
                                                    X_2*G.2, X_2*G.3)) )

## Remove R objects
rm(model, samp, coefs, model_string, n, X_3, X_4, X_5, betas)

```





<br>
<br>

#### Figure 6: The group-level mean functions from JAGS using a 5th-order linear model with time-by-group and $\text{Time}^2$-by-group interactions.
```{r, echo=FALSE}

## Create the plot
figure = ggplot(data = dat, aes(x = Time, y = Y_Smooth_Fitted_JAGS_5, group = ID, color = Group)) +
  geom_line() +
  theme_bw() +
  scale_color_brewer(palette = "Set1") +
  xlab("Time") +
  ylab("Y")

## Save the plot
ggsave(plot = figure,
       filename = "figure6.pdf",
       device = "pdf",
       path = "figs/",
       width = 8,
       height = 6)

## Return the plot
figure
rm(figure)

```





<br>
<br>
\pagebreak

## Analyze Using Penalized Splines and JAGS

Fit a penalized spline using JAGS:
```{r}

# Define values
n = nrow(dat)

## Create the model
model_string <- "model{

# Likelihood
for(i in 1:n){
Y[i]   ~ dnorm(mu[i], inv.var)
mu[i] <- beta[1] + beta[2]*G.2[i] + beta[3]*G.3[i] + beta[4]*X[i] + beta[5]*X_2[i] + alpha[1]*random.basis.1[i,1] + alpha[2]*random.basis.1[i,2] + alpha[3]*random.basis.1[i,3] + alpha[4]*random.basis.1[i,4] + alpha[5]*random.basis.1[i,5] + alpha[6]*random.basis.1[i,6] + alpha[7]*random.basis.1[i,7] + alpha[8]*random.basis.1[i,8] + alpha[9]*random.basis.1[i,9] + alpha[10]*random.basis.1[i,10] + alpha[11]*random.basis.1[i,11] + alpha[12]*random.basis.1[i,12] + alpha[13]*random.basis.1[i,13] + alpha[14]*random.basis.1[i,14] + alpha[15]*random.basis.1[i,15] + alpha[16]*random.basis.1[i,16] + alpha[17]*random.basis.1[i,17] + alpha[18]*random.basis.1[i,18] + alpha[19]*random.basis.1[i,19] + alpha[20]*random.basis.1[i,20] + alpha[21]*random.basis.1[i,21] + alpha[22]*random.basis.1[i,22] + alpha[23]*random.basis.1[i,23] + alpha[24]*random.basis.1[i,24] + alpha[25]*random.basis.2[i,1] + alpha[26]*random.basis.2[i,2] + alpha[27]*random.basis.2[i,3] + alpha[28]*random.basis.2[i,4] + alpha[29]*random.basis.2[i,5] + alpha[30]*random.basis.2[i,6] + alpha[31]*random.basis.2[i,7] + alpha[32]*random.basis.2[i,8] + alpha[33]*random.basis.2[i,9] + alpha[34]*random.basis.2[i,10] + alpha[35]*random.basis.2[i,11] + alpha[36]*random.basis.2[i,12] + alpha[37]*random.basis.2[i,13] + alpha[38]*random.basis.2[i,14] + alpha[39]*random.basis.2[i,15] + alpha[40]*random.basis.2[i,16] + alpha[41]*random.basis.2[i,17] + alpha[42]*random.basis.2[i,18] + alpha[43]*random.basis.2[i,19] + alpha[44]*random.basis.2[i,20] + alpha[45]*random.basis.2[i,21] + alpha[46]*random.basis.2[i,22] + alpha[47]*random.basis.2[i,23] + alpha[48]*random.basis.2[i,24] + alpha[49]*random.basis.3[i,1] + alpha[50]*random.basis.3[i,2] + alpha[51]*random.basis.3[i,3] + alpha[52]*random.basis.3[i,4] + alpha[53]*random.basis.3[i,5] + alpha[54]*random.basis.3[i,6] + alpha[55]*random.basis.3[i,7] + alpha[56]*random.basis.3[i,8] + alpha[57]*random.basis.3[i,9] + alpha[58]*random.basis.3[i,10] + alpha[59]*random.basis.3[i,11] + alpha[60]*random.basis.3[i,12] + alpha[61]*random.basis.3[i,13] + alpha[62]*random.basis.3[i,14] + alpha[63]*random.basis.3[i,15] + alpha[64]*random.basis.3[i,16] + alpha[65]*random.basis.3[i,17] + alpha[66]*random.basis.3[i,18] + alpha[67]*random.basis.3[i,19] + alpha[68]*random.basis.3[i,20] + alpha[69]*random.basis.3[i,21] + alpha[70]*random.basis.3[i,22] + alpha[71]*random.basis.3[i,23] + alpha[72]*random.basis.3[i,24]
}

# Prior for beta
for(j in 1:5){
beta[j] ~ dnorm(0,0.0001)
}

# Prior for alpha
for(k in 1:(3*K)){
alpha[k] ~ dnorm(0,0.0001)
}

# Prior for the inverse variance
inv.var    ~ dgamma(0.01, 0.01)
sigma     <- 1/sqrt(inv.var)

}"

## Compile the model
model = jags.model(textConnection(model_string),
                   data = list(Y = Y,
                               n = n,
                               X = X,
                               X_2 = X_2,
                               G.2 = G.2,
                               G.3 = G.3,
                               K = K,
                               random.basis.1 = random.basis.1,
                               random.basis.2 = random.basis.2,
                               random.basis.3 = random.basis.3),
                   n.chains = 3,
                   n.adapt  = 1000)

## Burn-in samples
update(model,
       1000,
       progress.bar = "none")

## Draw additional samples
samp = coda.samples(model,
                    variable.names = c("beta","alpha","sigma"),
                    thin = 3,
                    n.iter = 5000,
                    progress.bar = "none")

```





```{r, include=FALSE}

## Save the model objects
save(model,
     samp,
     file = "output/model_JAGS_spline.rda")

```





<br>
<br>

Investigate the fit of the model and extract the fitted values:
```{r}

## Investigate the model
summary(samp)
plot(samp)

## Save the coefficients from the model
coefs = data.frame( summary(samp)$statistics )

## Save the lambda estimates
#var.e4
#var.u4

## Extract the parameter estimates and order them c(fixed, random)
params = c( coefs$Mean[ row.names(coefs) %in% paste0( paste0("beta[",1:5),"]") ],
            coefs$Mean[ row.names(coefs) %in% paste0( paste0("alpha[",1:72),"]") ] )

## Calculate the fitted values
dat$Y_Smooth_Fitted_JAGS_Spline = c(t(params) %*% t(cbind(1,
                                                          G.2,
                                                          G.3,
                                                          X,
                                                          X_2,
                                                          random.basis.1,
                                                          random.basis.2,
                                                          random.basis.3)) )
rm(coefs, model, samp, model_string, random.basis.1, random.basis.2,
   random.basis.3, n, G.2, G.3, K, params)

```





<br>
<br>

#### Figure 7: The group-level mean functions using penalized splines and JAGS.
```{r, echo=FALSE}

## Create the plot
figure = ggplot(data = dat, aes(x = Time, y = Y_Smooth_Fitted_JAGS_Spline, group = ID, color = Group)) +
  geom_line() +
  theme_bw() +
  scale_color_brewer(palette = "Set1") +
  xlab("Time") +
  ylab("Y")

## Save the plot
ggsave(plot = figure,
       filename = "figure7.pdf",
       device = "pdf",
       path = "figs/",
       width = 8,
       height = 6)

## Return the plot
figure
rm(figure)

```





<br>
<br>
\pagebreak

#### Figure 8: A comparison of the fitted effect functions from each model.
```{r, echo=FALSE}

## Subset the data to only 1 ID per group
dat.temp = dat[dat$ID %in% c(1, 11, 21), ]

## Combine the fitted values
temp = data.frame(Y = c(dat.temp$Y_True,
                        dat.temp$Y_Smooth_Fitted_lm,
                        dat.temp$Y_Smooth_Fitted_lme,
                        dat.temp$Y_Smooth_Fitted_JAGS_5,
                        dat.temp$Y_Smooth_Fitted_JAGS_Spline),
                  X = rep(dat.temp$Time, 5),
                  Group = rep(dat.temp$Group, 5),
                  Model = c( rep("True Functions", nrow(dat.temp)),
                             rep("lm() Model", nrow(dat.temp)),
                             rep("lme() Spline", nrow(dat.temp)),
                             rep("JAGS Model", nrow(dat.temp)),
                             rep("JAGS Spline", nrow(dat.temp)) ) )
rm(dat.temp)

## Create factors
temp$Model = factor(temp$Model,
                    levels = c("True Functions","lm() Model","lme() Spline","JAGS Model","JAGS Spline"))
temp$Group = factor(temp$Group)

## Compare the fits
figure = ggplot(data = temp, aes(x = X, y = Y, color = Group)) +
  geom_line(aes(linetype = Model), size = 1.05) +
  theme_bw() +
  scale_color_brewer(palette = "Set1") +
  xlab("Time") +
  ylab("Y") +
  geom_point(data = dat, aes(x = Time, y = Y_Sim, color = Group), size = 0.5, alpha = 0.2)

## Save the plot
ggsave(plot = figure,
       filename = "figure8.pdf",
       device = "pdf",
       path = "figs/",
       width = 8,
       height = 6)

## Return the plot
figure
rm(figure, temp)

```





```{r, include=FALSE}

## Save the dataset with all of the fitted values
save(list = ls(),
     file = "data/finaldata.rda")

```





<br>
<br>
\pagebreak

## Session Info

A summary of the R session used for the analysis.
```{r, echo=FALSE}

print(sessionInfo(), local = FALSE)

```


